---
title: "Illustration of sicomore function to detect interactions between 2 datasets"
author: "Florent Guinot"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEncoding{UTF-8}
---
```{r message=FALSE, warning=FALSE, include=FALSE}
load("bench.R")
library(microbenchmark)
library(latex2exp)
library(ggplot2)
library(magrittr)
```
# Data sets generations

Let us first load the package plus some basic functions for data generation that can be found in the "inst" directory.

```{r, message=FALSE, warning=FALSE}
set.seed(1234)
library(SIComORe)
library(Matrix)
library(mvtnorm)

```

## Input Matrices generations

This goes pretty much just like in the example on an single data set (see the corresponding vignette), but two data set are required. The sample size is common to both data sets:

```{r}
n <- 100
```

Now, the first data set:
```{r}
## FIRST DATA
p1 <- 50 # number of variables 
K1 <- 5  # number of groups
rhos <- K1 %>% 
  runif(.5,.95) # correlation within groups
grp1.size <- rmultinom(1,p1,rep(p1/K1,K1)) ## group sizes
Sigma1 <- lapply(1:K1, function(k) return(matrix(rhos[k],grp1.size[k],grp1.size[k]))) %>% 
  bdiag()
diag(Sigma1) <- 1
grp1 <- rep(seq(grp1.size),grp1.size)
X1 <- Sigma1 %>% 
  as.matrix() %>% 
  rmvnorm(n, sigma=.) %>% 
  scale()
```

And the second data set:
```{r}
p2 <- 40
K2 <- 10
rhos <- K2 %>% 
  runif(.5,.95) # correlation within groups
grp2.size <- rmultinom(1,p2,rep(p2/K2,K2)) ## group sizes
Sigma2 <- lapply(1:K2, function(k) return(matrix(rhos[k],grp2.size[k],grp2.size[k]))) %>%
  bdiag()
diag(Sigma2) <- 1
grp2 <- rep(seq(grp2.size), grp2.size)
X2 <- Sigma2 %>% 
  as.matrix() %>% 
  rmvnorm(n, sigma=.) %>% 
  scale()
```

We show the matrix of empirical correlation exhibiting the block structure of both matrices:
```{r, fig.width=7}
par(mfrow = (c(2,1)))
image(Matrix(cor(X1)), main = "cor(X1)")
image(Matrix(cor(X2)), main = "cor(X2)")
```

### Phenotype/Response vector

Generation of the response vector and vector of coefficient is a little more tricky this time

First, we need first to compress each input matrices, 
```{r}
simple.effects.1 <- t(rowsum(t(X1), grp1)/tabulate(grp1))
simple.effects.2 <- t(rowsum(t(X2), grp2)/tabulate(grp2))
```

Then, we compute the corresponding matrix of data interaction 
```{r}
interactions <- lapply(1:ncol(simple.effects.2), function(i) sweep(simple.effects.1, 1, simple.effects.2[, i], "*")) %>%
  do.call(cbind,.)
```

The parameters in the linear model originating the response variable are chosen such that there is $s=2$ non-null interactions. The simple effect associated with these interactions are non-null too.

```{r}
s <- 2 ## number of non-null interactions
# vector of true parameter at the correct level of the hierarchies
dim1 <- ncol(simple.effects.1)
dim2 <- ncol(simple.effects.2)
dim.inter <- dim1 * dim2

ind.inter <- sample(1:dim.inter, s)
theta.inter <- rep(0, dim.inter)
theta.inter[ind.inter] <- runif(s, min=2, max=4) ## interactions : sparse
theta.block.inter <- Matrix(theta.inter, dim1, dim2)

ind.simple.effects <- which(theta.block.inter!=0, arr.ind=TRUE)
theta1 <- rep(0, dim1)
theta1[ind.simple.effects[,1]] <- runif(nrow(ind.simple.effects), min=1,max=2)

theta2 <- rep(0, dim2)
theta2[ind.simple.effects[,2]] <- runif(nrow(ind.simple.effects), min=1,max=2)
```

Finally we can draw the response vector
```{r}
sigma <- 3
epsilon <- rnorm(n) * sigma
mu <- 3
y <- mu + interactions %*% theta.inter + simple.effects.1 %*% theta1 + simple.effects.2 %*% theta2 + epsilon
r2 <- 1-sum(epsilon^2) / sum((y-mean(y))^2)
r2
```

## Adjusting multiple hierarchies

Let us run the main `sicomore` function using Ward's algorithm for the hierarchical clustering:

```{r message=FALSE, warning=FALSE, include=FALSE}
res <- sicomore(y, list(X1, X2), selection = "sicomore", choice=c("lambda.min", "lambda.min"), method.clus = c("ward.D2","ward.D2"))
```

We can plot the most striking interactions with the method `plot()` and compare with the "true" model: 

```{r message=FALSE, warning=FALSE, fig.width=7}
theta.full <- theta.block.inter[rep(1:nrow(theta.block.inter), table(grp1)), 
                                rep(1:ncol(theta.block.inter), table(grp2))] %>%
  as.matrix()

theta.true <- reshape2::melt(theta.full) %>% 
  ggplot(aes(Var1, Var2, fill=value)) +
  geom_tile(show.legend=FALSE, colour = "grey80") +
  scale_fill_gradient2(high = "grey40") +
  theme_minimal() +
  ggtitle("Matrix of true interactions") +
  labs(x = TeX('Variables in $\\mathbf{X}_1$'), y = TeX('Variables in $\\mathbf{X}_2$$')) +
  coord_fixed()
theta.estimated <- res$plot(main="Estimated matrix of interactions", threshold=.05)
cowplot::plot_grid(theta.estimated,theta.true, ncol=2)
```

## Comparison of available options for variable selection
To compare the 3 possible options "mlgl","hcar and "sicomore" in `selection`, we consider the performance in term of variable selection. So let us first define an simple function to compute basic statistics:

```{r}
getSelPerf <- function(theta, theta.star) {
  ones.true <- which(theta.star != 0)
  zero.true <- which(theta.star == 0)

  ones.hat <- which(theta != 0)
  zero.hat <- which(theta == 0)

  tp <- sum(ones.hat %in% ones.true)
  tn <- sum(zero.hat %in% zero.true)

  fp <- sum(ones.hat %in% zero.true)
  fn <- sum(zero.hat %in% ones.true)

  if ((tp+fn)==0) recall    <- 0 else recall    <- tp/(tp+fn) 
  if ((fp+tn)==0) fallout   <- 0 else fallout   <- fp/(fp+tn) 
  if ((tp+fp)==0) precision <- 0 else precision <- tp/(tp+fp) 

  return(c(recall=recall, fallout=fallout, precision=precision))
}
```

And let us compare the 3 options in `selection`:

```{r message=FALSE, warning=FALSE}
out.sicomore <- sicomore(y, list(X1, X2), selection="sicomore", verbose = FALSE)
out.mlgl <- sicomore(y, list(X1, X2), selection="mlgl", verbose = FALSE)
out.hcar <- sicomore(y, list(X1, X2), selection="hcar", verbose = FALSE)
```  

We compute the matrix of interactions effects with the method `getSignificance(effect = "theta")` for each option:
```{r message=FALSE, warning=FALSE}
theta.mlgl <- out.mlgl$getSignificance(effect = "theta")
theta.hcar <- out.hcar$getSignificance(effect = "theta")
theta.sicomore <- out.sicomore$getSignificance(effect = "theta")
```

Finally we evaluate the ability to recover the true interaction term for several significance thresholds:
```{r message=FALSE, warning=FALSE, fig.width=7}
seq.thres <- seq(0,1,len=100)
res.sicomore <- sapply(seq.thres, function(threshold) getSelPerf(theta.sicomore>=threshold, theta.full)) %>%
  t() %>%
  data.frame()
res.mlgl <- sapply(seq.thres, function(threshold) getSelPerf(theta.mlgl>=threshold, theta.full)) %>%
  t() %>%
  data.frame()
res.hcar <- sapply(seq.thres, function(threshold) getSelPerf(theta.hcar>=threshold, theta.full)) %>%
  t() %>%
  data.frame()

res.df <- data.frame(method = rep(c("sicomore","mlgl","hcar"), each = 100), rbind(res.sicomore, res.mlgl, res.hcar))

res.df %>% 
  ggplot(aes(x = fallout, y = recall, color = method)) + 
  geom_line()
```

## Comparison of computation time between the 3 options:

```{r eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
SICOMORE <- expression(sicomore(y, list(X1, X2), selection="sicomore", verbose=FALSE))
MLGL <- expression(sicomore(y, list(X1, X2), selection="mlgl", verbose=FALSE))
HCAR <- expression(sicomore(y, list(X1, X2), selection="hcar", verbose=FALSE))
bench <- microbenchmark(eval(SICOMORE) , eval(MLGL), eval(HCAR), times = 50)
```
```{r,fig.width=7}
autoplot(bench)
```

