---
title: "Illustration of sicomore on two interaction data sets"
author: "Julien Chiquet"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEncoding{UTF-8}
---

# Data sets generations

Let us first load the package plus some basic functions for data generation that can be found in the "inst" directory.

```{r, message=FALSE, warning=FALSE}
set.seed(1234)
library(SIComORe)
library(Matrix)
library(mvtnorm)
```

## Input Matrices generations

This goes pretty much just like in the example on an single data set (see the corresponding vignette), but two data set are required. The sample size is common to both data sets:

```{r}
n <- 100
```

Now, the first data set:
```{r}
## FIRST DATA
p1 <- 50 # number of variables 
K1 <- 5  # number of groups
rhos <- runif(K1,.5,.95) # correlation within groups
grp1.size <- rmultinom(1,p1,rep(p1/K1,K1)) ## group sizes
Sigma1 <- bdiag(lapply(1:K1, function(k) return(matrix(rhos[k],grp1.size[k],grp1.size[k]))))
diag(Sigma1) <- 1
grp1 <- rep(1:length(grp1.size), grp1.size)
X1 <- scale(rmvnorm(n, sigma=as.matrix(Sigma1)))
```

And the second data set:
```{r}
p2 <- 40
K2 <- 10
rhos <- runif(K2,.5,.95) # correlation within groups
grp2.size <- rmultinom(1,p2,rep(p2/K2,K2)) ## group sizes
Sigma2 <- bdiag(lapply(1:K2, function(k) return(matrix(rhos[k],grp2.size[k],grp2.size[k]))))
diag(Sigma2) <- 1
grp2 <- rep(1:length(grp2.size), grp2.size)
X2 <- scale(rmvnorm(n, sigma=as.matrix(Sigma2)))
```

### Phenotype/Response vector

Generation of the response vector and vector of coefficient is a little more tricky this time

First, we need first to compress each input matrices, 
```{r}
simple.effects.1 <- t(rowsum(t(X1), grp1)/tabulate(grp1))
simple.effects.2 <- t(rowsum(t(X2), grp2)/tabulate(grp2))
```

Then, we compute the corresponding matrix of data interaction 
```{r}
interactions <- do.call(cbind, lapply(1:ncol(simple.effects.2), 
      function(i) sweep(simple.effects.1, 1, simple.effects.2[, i], "*")))
```

The parameters in the linear model oroginating the reponse are chosen such that there is $s=2$ non-null interactions. The simple effect associated with these interactions are non-null too.

```{r}
s <- 2 ## number of non-null interactions
# vector of true parameter at the correct level of the hierarchies
dim1 <- ncol(simple.effects.1)
dim2 <- ncol(simple.effects.2)
dim.inter <- dim1 * dim2

ind.inter <- sample(1:dim.inter, s)
theta.inter <- rep(0, dim.inter)
theta.inter[ind.inter] <- runif(s, min=2, max=4) ## interactions : sparse
theta.block.inter <- Matrix(theta.inter, dim1, dim2)

ind.simple.effects <- which(theta.block.inter!=0, arr.ind=TRUE)
theta1 <- rep(0, dim1)
theta1[ind.simple.effects[,1]] <- runif(nrow(ind.simple.effects), min=1,max=2)

theta2 <- rep(0, dim2)
theta2[ind.simple.effects[,2]] <- runif(nrow(ind.simple.effects), min=1,max=2)
```

Finally we can draw the response vector
```{r}
sigma <- 3
epsilon <- rnorm(n) * sigma
mu <- 3
y <- mu + interactions %*% theta.inter + simple.effects.1 %*% theta1 + simple.effects.2 %*% theta2 + epsilon
r2 <- 1-sum(epsilon^2) / sum((y-mean(y))^2)
r2
```

## Adjusting multiple hierarchies

Let us run the main `sicomore` function

```{r}
res <- sicomore(y, list(X1, X2))
```

We can plot the most striking interactions and compare with the "true" model: 

```{r}
theta.full <- as.matrix(theta.block.inter[rep(1:nrow(theta.block.inter), table(grp1)),
                                  rep(1:ncol(theta.block.inter), table(grp2))])

par(mfrow=c(1,2))
res$plot(main="estimated")
image(theta.full, col=terrain.colors(3, alpha = 0.4),main="TRUE",axes=FALSE)
title(outer=TRUE, main="\nPerformance for interaction recovery", )
```

To compare different variant, we consider the performance in terme of variable selection. So let us first define an simple function to compute basic statistics:

```{r}
getSelPerf <- function(theta, theta.star) {
  ones.true <- which(theta.star != 0)
  zero.true <- which(theta.star == 0)

  ones.hat <- which(theta != 0)
  zero.hat <- which(theta == 0)

  tp <- sum(ones.hat %in% ones.true)
  tn <- sum(zero.hat %in% zero.true)

  fp <- sum(ones.hat %in% zero.true)
  fn <- sum(zero.hat %in% ones.true)

  if ((tp+fn)==0) recall    <- 0 else recall    <- tp/(tp+fn) 
  if ((fp+tn)==0) fallout   <- 0 else fallout   <- fp/(fp+tn) 
  if ((tp+fp)==0) precision <- 0 else precision <- tp/(tp+fp) 

  return(c(recall=recall, fallout=fallout, precision=precision))
}
```

And let us compare the 3 variants

```{r}
fast <- sicomore(y, list(X1, X2), selection="fast")
mlgl <- sicomore(y, list(X1, X2), selection="mlgl")
hcar <- sicomore(y, list(X1, X2), selection="hcar")
  
theta.mlgl <- mlgl$getSignificance()
theta.hcar <- hcar$getSignificance()
theta.fast <- fast$getSignificance()

seq.thres <- seq(0,1,len=100)
res.fast <- data.frame(t(sapply(seq.thres, function(threshold) getSelPerf(theta.fast>=threshold, theta.full))))
res.mlgl <- data.frame(t(sapply(seq.thres, function(threshold) getSelPerf(theta.mlgl>=threshold, theta.full))))
res.hcar <- data.frame(t(sapply(seq.thres, function(threshold) getSelPerf(theta.hcar>=threshold, theta.full))))

par(mfrow=c(1,1))
plot(res.fast$fallout, res.fast$recall, type="l")
lines(res.mlgl$fallout, res.mlgl$recall,col=2)
lines(res.hcar$fallout, res.hcar$recall,col=3)
legend("bottomright", legend = c("fast", "MLGL", "HCAR"), col=1:3, lty=1)
```

